### 2025-11-14 Paper/paperdraft_fall2025.tex
Updates from Overleaf

	\linespread{1.2}\title{\vspace{-0.5in} Does Hospital Leadership Matter?\\ \large Evidence from Pay-for-Performance}

@@

    \begin{abstract}
		{\small
         Firm managers are often seen as key drivers of firm decisions, strategies, and performance. While extensive research links managerial characteristics to firm outcomes, less is known about how hospital executives influence responses to health care incentives. Health care policies in the US often focus on improving quality of care without increasing costs, yet it is unclear whether hospital leaders might contribute to this goal. Leveraging pay-for-performance incentives, I estimate the differential response to financial incentives on quality by different types of nonprofit hospital executives: those with and without a clinical training background as a medical doctor. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. I find that this difference can be explained by patient-centered focus and by expertise lowering the cost of quality improvement. These findings highlight the importance of considering leadership as an important factor in hospital decision making when adjusting incentives. 
		} 
	\end{abstract}

@@

    In corporate environments, executives are considered crucial drivers of a company's practices and performance, as they are the highest level of management that affects day-to-day operations of the organization. Executive heterogeneity has been linked to various actions and outcomes of firms; characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, our understanding is typically limited to executives working for large, publicly traded corporations. Additionally, while correlations are an important first step to understanding the effects of leadership, little work has been done to uncover whether such correlations occur due to executive characteristics interacting with the objectives of firms.

    In this paper, I study the effect of specialized executive training on firm behavior when the firm experiences an external shock to their objective function, specifically in the context of US hospitals. I leverage this setting for two reasons. First, in the US, health care makes up almost 20\% of spending (\cite{CMSNationalHealthExpenditure}), and a major goal of policymakers is to improve the quality of health care without increasing costs, as the US preforms poorly in these metrics compared to other similar countries. Yet, little is known about how leaders affect health care value in the hospitals they run. Second, nonprofit hospitals are particularly important in this context, and it is unclear whether existing findings on executive characteristics can be extrapolated to firms with presumably different incentives and objectives than those which have been previously studied.

    An executive characteristic relevant to this context is clinical training. Physicians can bring a unique combination of clinical and administration expertise to leadership, and they have insights that can potentially benefit their employees and improve quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how the presence of clinically trained executives on a hospital leadership team affects how the hospital makes decisions. I scrape executive names and titles of nonprofit hospitals in the US from publicly available Tax Form 990s. Combining this with other public hospital data, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is an important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

    The non-random selection of leaders makes it difficult to interpret any direct comparison of differing executive teams as causal. Therefore, I leverage pay-for-performance policies enacted in the US as an exogenous shock to hospital incentives on quality, and compare the change in readmission and mortality rates of hospitals with and without clinically trained executives. Under the assumption that the two types of teams would have had similar trends in readmission and mortality rates absent the incentive change, I find that hospitals without clinically trained trained executives respond more to financial incentives on quality than hospitals with clinical executives. That is, while both types of hospital leadership teams decrease readmission rates after the policy change, those without clinical executives decrease readmissions at a faster rate than those with clinical executives. I find no evidence of selective patient practices driving this result.

    I use a theoretical model to investigate whether these results might be explained by differing preferences on the trade-off between profit and patient welfare, where I model these hospital objectives and allow the weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing the response of hospitals with and without clinically trained executives to the response of for-profit hospitals, who have revealed that they are more profit driven by their ownership type. I find that nonprofit hospitals without clinically trained executives respond to the financial incentives on quality similarly to for-profits, while nonprofits without clinically trained executives respond differently than for-profits. For-profits and non-clinical executive teams change quality more in response to the incentive change than clinically trained executive teams.

    Finally, the question remains as to whether clinically trained executives are simply a signal of unchanging underlying hospital preferences (of, for example, the board of directors), or if they actively manage the hospital differently, effectively changing hospital objectives. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. To measure the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by clinically trained executives managing the hospital differently.

    Many researchers have documented a correlation between executive characteristics and firm actions, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Having more female executives is correlated with female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}; \cite{adams2018director}; \cite{frydman2019rising}); older CEOs are associated with measures of corporate culture (\cite{graham2022corporate}). Finally, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}).

    Due to data limitations, many of these studies focus on large, publicly traded firms in the US. \citeauthor{brickley2010board} (\citeyear{brickley2010board}) have the only study to my knowledge in the US nonprofit context considering hospital board of directors. Using exogenous variation in expected Medicare profits, the authors find that having an internal board of directors increases CEO compensation, and having physician board members decreases public donations (\cite{brickley2010board}). Additionally, two studies focus on hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) use data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{otero2022managers} (\citeyear{otero2022managers}) investigate the role of CEOs in public hospitals in Chile, and finds an 8\% decrease in mortality rates for hospitals with top managers. Further, a major contributor of this quality improvement is the exodus of older physicians as CEOs (\cite{otero2022managers}). Finally, while not in the context of hospitals, \citeauthor{la2023impact} (\citeyear{la2023impact}) finds that OB-GYNs acquired by financially driven Physician Practice Management Companies (PPMCs) increase C-Sections compared to those acquired by clinically driven PPMCs (\cite{la2023impact}). Thus, the motivations of the management of health care facilities matters for provider behavior. I contribute to our knowledge of how executive team characteristics affect firm behaviors by (1) considering the important setting of nonprofit hospitals, which serve the majority of patients in the US, (2) considering a unique and relevant executive characteristic of the entire executive team, and (3) observing firm behavior in the presence of an exogenous incentive change.

    Additionally, I contribute to our understanding of how providers respond to pay-for-performance incentives in health care. The most in depth study of how hospitals respond to the Hospital Readmissions Reduction Program, one of the two large pay-for-performance initiatives, is \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). This study finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Around 40\% of the decrease in readmissions is due to selective patient practices. Further, market concentration and hospital systems play a large role in how hospitals respond to the program (\cite{kunz2024assessing}). Research on the other large pay-for-performance initiative, the Hospital Value Based Purchasing Program, generally finds that the program had no effect on underlying hospital quality (\cite{us2015hospital}; \cite{norton2018moneyball}; \cite{friedson2019so}). This paper contributes to our understanding of how characteristics of hospitals affect response to pay-for-performance incentives.

@@

    I examine three outcomes to test this hypothesis.First, I use the CMS Case Mix Index, which provides an aggregate measure of patient complexity for each hospital and year. A higher index indicates that the hospital’s patient population requires more complex treatment. Additionally, I examine the share of patients treated in the hospital who have public insurance, either Medicare or Medicaid. These outcomes help assess whether certain leadership types attempt to shift costs relative to others after penalties or rewards take effect.

    I estimate the same synthetic difference-in-differences specification described in Section \ref{sec:clinical}. The results are presented in Table \ref{tab:selection}. All estimated treatment effects are statistically insignificant and near zero, indicating no evidence that overall hospital patient populations change differentially by leadership type. This does not rule out changes in specific readmission behavior, as documented by \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). However, a limitation of this dataset is that it only permits analysis of aggregate patient populations rather than individual readmissions.

@@

    While previous literature has documented correlations between executive characteristics and firm behaviors/performance, research on this topic remains limited in the health care sector. Health care managers are tasked with maintaining a business while also prioritizing patient care, which is why existing findings on manager effects may not extend to this setting. This paper investigates whether having executive leaders with clinical training impacts hospital behaviors and outcomes, particularly in response to a policy change that financially incentivizes quality. I utilize a unique data set of nonprofit US hospitals that captures the proportion of clinical executives, specialties of clinical executives, hospital characteristics, and hospital quality metric outcomes.

    In 2012, two pay-for-performance policies went into effect that either penalized hospitals with high readmission rates, or gave payments to hospitals with low readmission/mortality rates. These policies were relevant to hospital finances, and gave tangible incentive for hospitals to increase quality. Thus, I leverage this exogenous policy change and estimate how clinically trained executives affect readmission and mortality rates in response to the change using a synthetic difference-in-differences strategy. I find that hospitals without clinically trained executives improve quality more than hospitals without clinically trained executives after the incentive change.

    I investigate several hypotheses that arise given the estimated difference in response. First, this result is not driven by non-clinical teams adjusting their patient complexity. However, comparing non-clinical teams to for-profit hospitals reveals that it could be underlying preferences for profit vs. patient benefit that drive the difference. Finally, I leverage exogenous changes in executive teams to show that clinically trained executives are not simply a signal of underlying preferences, but manage the hospital differently.

    In US health care, a very common policy goal is to improve quality of care without increasing costs. However, up to this point, it was unclear whether hospital leaders might play a role in attaining this. In this paper, I show that having hospital managers with clinical training does affect the way hospitals respond to incentives. Importantly, my results point to clinically trained executives already prioritizing quality before incentives were put in place. My findings have implications for policy makers considering financial incentives on quality, and even point to the potential of incentivizing the presence of clinically trained hospital leaders.

@@

\appendix

@@

I use the API of ProPublica's Nonprofit Explorer to access the archive of nonprofit Tax Form 990s.\footnote{At the time of writing this, information on using version 2 of the API can be found at \hyperlink{https://projects.propublica.org/not-for-profits/api}{https://projects.propublica.org/not-for-profits/api}.} I extract the Employee Identification Number (EIN) for all nonprofits categorized with a National Taxonomy of Exempt Entities (NTEE) code of E20 (hospital), E21 (community health system), and E22 (general hospital). There are 5,588 EINs total in this list.

I query more detailed information from the API on each hospital EIN. I save the name, secondary name, state, and zip code, all of which do not vary by year. I also record URL links to the Tax Form 990 PDFs, which are unique in each year. For the sake of a comprehensive data set, I keep years 2006-2020 (I later limit to 2010-2014 when focusing on pay-for-performance initiatives in 2012). Thus, I finish this step with a panel data set of EIN characteristics and PDF locations. Importantly, there are multiple types of Tax Form 990s depending on the size of the nonprofit. In many cases, one nonprofit has at least two different forms filed in a given year. I filter out any EIN-years for which there are no PDFs available. There are numerous foundations or auxiliary firms with the purpose of raising funds for the hospital, but do not provide services to patients. I filter out any not-for-profit with ``foundation" or ``auxiliary" in the name. I also filter out various specialty centers such as hospice or cancer centers.

@@

I proceed matching based on names in several steps. I focus on exact string matches, so I remove all spaces and common characters that could cause mismatches such as \&, ', -, and ``inc". I loop through each AHA ID, limit to nonprofits in the same state, and extract any exact name matches. When an exact match is found, I record the link between AHA ID and EIN. In this first layer of matching, 860 hospitals in the AHA data are linked to an EIN, equivalent to 31\% of nonprofit hospitals in the sample. In the next layer of matching, I remove common words such as ``healthcare", ``regional", ``hospital", etc. That way if there are subtle differences in names, removing common words may allow for an exact match. Again, I take each AHA hospital name and look for exact matches in the nonprofits in the same state. This adds an additional 90 hospital matches, accounting for a total of 34.5\% of AHA hospitals. Finally, I manually search through unmatched EINs to identify any matches. From google searches, I identify an additional 300 EIN-AHA ID matches. yielding a total of 1200 EINs.

While it would be ideal to analyze all nonprofit hospitals, a sample size of 852 firms is relatively large in the hospital executive literature. I present a comparison of in-sample and out-of-sample nonprofits. I present means and standard deviations of these measures in Table \ref{tab:nfp_sample_compare}. The in-sample nonprofits are slightly larger in terms of beds and patients seen. They are also slightly more likely to be penalized or receive payments under the pay-for-performance incentives. However, average readmission and mortality rates are very similar, as well as case mix index. The largest difference in the two samples is that I under-represent nonprofit hospitals in systems. This is due to the nature of the tax form 990s, where systems often file one tax form for the entire system, making it difficult to discern the true managers of a specific hospital. Because of this, I drop hospitals from the sample who only have leadership information from a system tax form.

@@

I then extract the names of board members and executives from the Tax Form 990 PDFs of matched hospitals. In the data set of hospital PDF URLs that I collected earlier, I limit to the hospitals with solid matches described above. I then loop through each EIN, downloading PDFs locally and using the tesseract package in R to extract text from the relevant pages of the PDF using OCR text extraction methods. In particular, I loop through each page of the PDF, look for the title associated with leadership names: ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees", and save all the text from any pages where this title is found. I save the text to a list of all EIN, years present.

One aspect of the NonProfit Explorer API is that, only in some cases, if two forms are present for an EIN, year, only the first one (which is typically not the one with the relevant information) is pulled. Therefore, for some hospitals, a couple years will have gaps in text extraction data. I locate EIN, years where this problem is occurring, and a team of RAs locates and downloads the correct forms manually. I extract text from these manually downloaded forms in the same manner as above.

@@

The data is now an executive-year level data set with information on clinical training and specialty when relevant. I then create hospital level indicators based on the names, titles and positions associated with the hospital: the number of clinically trained executives, the number of executives with particular specialties, the number of total executives, and whether the hospital employs a CMO.

@@

For the main analysis, I limit to hospitals who either always have a clinically trained executive or never have a clinically trained executive. In a supplemental analyses, I leverage changes to hospital propensity to hire a clinically trained executive. I explore the exogeneity of such changes in Section \ref{sec:changes}. Additionally, I present the number of hospitals who change their propensity to hire a clinically trained executive at different times in Table \ref{tab:change_timing}. I break the timing up into pre-2012, 2012, and post-2012. I show the number of hospitals who have clinically trained executives in different time combinations. The number of hospitals that only have an MD executive in 2012, but at no other time, is only 6. The majority of hospitals (187) who ever have a clinical executive have one in every time period.

@@

@@

    \subsection{Binned Treatment}\label{app:binned}

@@

    \subsection{Selective Patient Practices}\label{app:casemix}

    A method hospitals have been shown to employ in order to decrease readmission rates is being more selective with which patients to admit to the hospital (\cite{gupta2021impacts}), as only admitting healthy patients mechanically decreases readmission rates. One hypothesis behind a more drastic quality decrease by non-clinical teams is that they may be more willing to participate in such practices, while executives with clinical training are less likely to turn patients away.

    To test this, I employ case mix index as the outcome in specification \ref{eq:clinical} to determine whether one type of hospital differentially changes patient composition after pay-for-performance incentives, where the same identification assumptions hold. I present the estimated difference in Figure \ref{fig:main_cmi_clinical}, comparing hospitals that never have a clinical executive to hospitals that always have a clinical executive. Neither type of hospital drastically changes case mix index after the incentive change, and there is no estimated differential response in complexity of patients seen in the hospital. Thus, the estimated difference in readmission rate response between clinical and non-clinical executives is not due to differentially changing case mix.

    \begin{figure}[ht!]
        \centering
        \caption{Effect of Clinical Executive on Case Mix Index}
        \includegraphics[width=0.5\textwidth]{Objects/cmi_md_nomd_synth_graph.pdf}
        \label{fig:main_cmi_clinical}
    \end{figure}

>>>>

	\linespread{1.2}\title{\vspace{-0.5in} DOES HOSPITAL LEADERSHIP MATTER?\\ \large EVIDENCE FROM PAY-FOR-PERFORMANCE\thanks{I thank Aliyah Bagro and Vidyah Vedula for their excellent research assistance. I express my deepest thanks to Ian McCarthy for his advice and ongoing helpful feedback. Further, I thank Sara Markowitz, Matt Grennan, and Evan Saltzman for their helpful comments. I am grateful to the audiences at EuHEA and ASHEcon conferences, as well as at the University of Queensland, the University of Melbourne, the University of Miami, and the University of Iowa for their insightful feedback.}}

@@

    \begin{abstract}
		{\small
         Firm managers are often seen as key drivers of firm decisions, strategies, and performance. Extensive research links managerial traits to firm outcomes, yet little is known about how hospital executives shape responses to healthcare incentives. Health care policies in the US often focus on improving quality of care without increasing costs, yet it is unclear whether hospital leaders might contribute to this goal. Leveraging pay-for-performance incentives, I estimate the differential response to financial incentives on quality by different types of nonprofit hospital executives: those with and without training as a medical doctor. I find that hospitals led by non-clinical executives respond more strongly to financial incentives than those led by clinically trained executives. The difference is not a signal of underlying hospital objectives, but a difference in active management. Further, the observed difference reflects two mechanisms: clinically trained leaders prioritize patient-centered care and possess expertise that reduces the cost of quality improvements. These findings highlight the role of leadership in firm decision-making and the need to account for executive characteristics when designing incentives.
		} 
	\end{abstract}

@@

    In corporate environments, executives are key drivers of company practices and performance, shaping day-to-day operations at the highest level of management. Executive heterogeneity (based on gender, age, and professional background) has been linked to firm outcomes, including financial performance and behaviors such as workforce diversity, wage setting, and merger strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, most research focuses on executives in large, publicly traded corporations. Moreover, while correlations provide an initial understanding of leadership effects, few studies examine whether these patterns arise from interactions between executive characteristics and firm objectives.

    In this paper, I study the effect of specialized executive training on firm behavior when an external shock alters the objective function, specifically in the context of US hospitals. I focus on this setting for two reasons. First, in the US, health care makes up almost 20\% of spending (\cite{CMSNationalHealthExpenditure}), and a major goal of policymakers is to improve the quality of health care without increasing costs, as the US performs poorly in these metrics compared to other high-income countries. Yet, little evidence exists on how hospital leaders affect health care value. Second, nonprofit hospitals are particularly important, and it is unclear whether existing findings on executive characteristics from other sectors can be extrapolated to firms with presumably different incentives and objectives.

    A key executive characteristic in context is clinical training. Physicians combine clinical and administrative expertise, offering insights that may benefit staff and improve care quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial because many lack formal business training, which can result in inefficient management and reduced quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how the presence of clinically trained executives on a hospital leadership team influences hospital decision making. I scrape executive names and titles of nonprofit hospitals in the US from publicly available Tax Form 990s. Combining this with other public hospital data, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is an important driver of health care in the US. Private nonprofit hospitals account for 50\% of all US hospitals and average 207 beds, compared to for-profit hospitals, which comprise 36\% and average 107 beds (\cite{ASPE_2023}).

    The non-random selection of leaders complicates causal interpretation of direct comparisons between executive teams. Therefore, I leverage pay-for-performance policies enacted in the US as an exogenous shock to hospital incentives on quality and compare changes in readmission and mortality rates between hospitals with and without clinically trained executives using a synthetic difference in differences estimation strategy. Assuming the two types of teams would have followed similar trends absent the policy change, I find that hospitals without clinically trained executives respond more strongly to financial incentives than those with clinical executives. In other words, both types of leadership teams reduce readmission rates after the policy change, but hospitals without clinical executives do so at a faster rate.

    I investigate several mechanisms to explain the observed difference. First, I leverage the timing of clinically trained executives to show that clinicians actively manage hospitals differently, rather than serving only as a signal of underlying preferences. Next, I develop a theoretical model to illustrate how these results may stem from differing preferences over the trade-off between profit and patient welfare, allowing the weight on each objective to vary by executive team type. Hospitals that prioritize profit respond more strongly to policies that incorporate quality into the profit function. Thus, one potential explanation is that non-clinical executive teams are more profit-driven. I test this empirically by comparing responses of hospitals with and without clinically trained executives to those of for-profit hospitals, which are known to be more profit-driven by ownership structure. I find that nonprofit hospitals without clinically trained executives respond to quality incentives similarly to for-profits, whereas those with clinically trained executives respond differently. Both for-profits and non-clinical executive teams increase quality more in response to the incentive change than clinically trained teams. Similarly, I show theoretically that leadership expertise may drive the observed difference and investigate this empirically by comparing physician specialties. This analysis suggests that expertise is a likely underlying mechanism. Finally, I find that changes in aggregate patient populations and investment decisions do not explain the observed difference.

    Many researchers have documented a correlation between executive characteristics and firm actions, beginning with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who shows that manager fixed effects are an important factor in many firm behaviors and decisions. Greater female representation among executives correlates with higher female employee wages and shifts in corporate strategy (\cite{flabbi2019female}; \cite{matsa2013female}). Young male CEOs tend to be more aggressive in mergers and acquisitions, whereas those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}). CEOs with general managerial ability receive higher pay and achieve better performance (\cite{kaplan2012ceo}; \cite{custodio2013generalists}; \cite{adams2018director}; \cite{frydman2019rising}). Older CEOs are associated with differences in corporate culture (\cite{graham2022corporate}). Finally, Chief Diversity Officers have not been found to influence diversity in university hiring (\cite{bradley2022impact}).

    Because of data limitations, many of these studies focus on large, publicly traded firms. To my knowledge, \citeauthor{brickley2010board} (\citeyear{brickley2010board}) provide the only U.S. nonprofit study examining hospital boards of directors. Using exogenous variation in expected Medicare profits, the authors find that internal boards increase CEO compensation, while physician board members reduce public donations (\cite{brickley2010board}). Two additional studies examine hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) use data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{munoz2025managers} (\citeyear{munoz2025managers}) investigate the role of CEOs in public hospitals in Chile, and find an 8\% decrease in mortality rates for hospitals with top managers. Further, a major contributor of this quality improvement is the exodus of older physicians as CEOs (\cite{munoz2025managers}). Finally, although not in a hospital context, \citeauthor{la2023impact} (\citeyear{la2023impact}) find that OB-GYNs acquired by financially driven Physician Practice Management Companies (PPMCs) perform more C-sections than those acquired by clinically driven PPMCs (\cite{la2023impact}). Thus, the motivations of the management of health care facilities matter for provider behavior. This paper contributes to understanding how executive team characteristics affect firm behavior by (1) examining the critical setting of nonprofit hospitals, which serve a large share of US patients, (2) analyzing a unique and relevant characteristic of the entire executive team, and (3) observing behavior under an exogenous incentive change.

    Additionally, this paper contributes to understanding how providers respond to pay-for-performance incentives in health care. The most in-depth study of hospital responses to the Hospital Readmissions Reduction Program (HRRP), one of the two major pay-for-performance initiatives, is by \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). This study finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Around 40\% of the decrease in readmissions is due to selective patient practices. Further, market concentration and hospital systems play a large role in how hospitals respond to the program (\cite{kunz2024assessing}). Research on the other large pay-for-performance initiative, the Hospital Value Based Purchasing Program, generally concludes that the program had no effect on underlying hospital quality (\cite{us2015hospital}; \cite{norton2018moneyball}; \cite{friedson2019so}). This paper extends this literature by examining how hospital leadership characteristics shape responses to pay-for-performance incentives.

@@

    I examine three outcomes to test this hypothesis. First, I use the CMS Case Mix Index, which provides an aggregate measure of patient complexity for each hospital and year. A higher index indicates that the hospital’s patient population requires more complex treatment. Additionally, I examine the share of patients discharged from the hospital who have public insurance, either Medicare or Medicaid. These outcomes help assess whether certain leadership types attempt to shift costs relative to others after penalties or rewards take effect.

    I estimate the same synthetic difference-in-differences specification described in Section \ref{sec:clinical}. The results are presented in Table \ref{tab:selection}. For Case Mix Index, clinically the estimate of -0.03 is statistically significant, but this is such a small magnitude (relative to a mean of 1.5) that this is a precise null effect. The estimated treatment effects for both Medicare and Medicaid populations are statistically insignificant and near zero, indicating no evidence that overall hospital patient populations change differentially by leadership type. This does not rule out changes in specific readmission behavior, as documented by \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). However, a limitation of this dataset is that it only permits analysis of aggregate patient populations rather than individual readmissions.

@@

    While previous literature has documented correlations between executive characteristics and firm behaviors/performance, research on this topic remains limited in the health care sector. Health care managers must balance business operations with patient care, so existing findings on manager effects may not fully apply to this setting. This paper examines whether executive leaders with clinical training influence hospital behaviors and outcomes, particularly in response to a policy change that financially incentivizes quality. I use a unique dataset of US nonprofit hospitals that includes the proportion and specialties of clinical executives, hospital characteristics, and quality outcomes.

    In 2012, two pay-for-performance policies went into effect: one penalized hospitals with high readmission rates, and the other rewarded hospitals with low readmission or mortality rates. These policies significantly affected hospital finances and created strong incentives to improve quality. I leverage this exogenous policy change to estimate how clinically trained executives influence readmission and mortality rates using a synthetic difference-in-differences approach. I find that hospitals without clinically trained executives improve quality more than hospitals with clinically trained executives after the incentive change.

    I explore several hypotheses to explain this differential response. First, I leverage the timing of clinically trained executive appointments to decompose the effect into signaling versus management. This decomposition shows that having a clinically trained executive in a leadership role during policy implementation is crucial, suggesting the difference reflects management practices rather than underlying hospital preferences. Next, I compare the two types of nonprofits to for-profit hospitals on average, revealing that non-clinical executive teams behave similarly to for-profits. I also examine whether the observed difference reflects the expertise of specific physician specialties by comparing internal medicine executives to others, and find evidence that this may be a contributing mechanism. Finally, I show that the difference is not explained by changes in aggregate patient populations or investment strategies.

    In US health care, a central policy goal is to improve quality of care without increasing costs. However, up to this point, it has been unclear whether hospital leaders might play a role in attaining this. The implications of the findings in this paper are two-fold. First, the background of hospital leaders in the US is an important factor for hospital behaviors, particularly whether the extent to which they respond to financial incentives. Policymakers should consider whether trends in clinician-led hospitals are shifting, as this will influence responses to policy incentives. Second, although the nonprofit versus for-profit debate remains important, leadership composition may be a more critical factor. Ultimately, understanding how leadership expertise shapes organizational responses to policy can inform more effective strategies for aligning financial incentives with quality goals.

@@

\appendix
\setcounter{table}{0}
\renewcommand{\thetable}{A.\arabic{table}}
\renewcommand{\thefigure}{A.\arabic{figure}}

@@

I use the API of ProPublica's Nonprofit Explorer to access the archive of nonprofit Tax Form 990s.\footnote{At the time of writing this, information on using version 2 of the API can be found at \href{https://projects.propublica.org/not-for-profits/api}{https://projects.propublica.org/not-for-profits/api}.} I extract the Employee Identification Number (EIN) for all nonprofits categorized under a National Taxonomy of Exempt Entities (NTEE) code of E20 (hospital), E21 (community health system), and E22 (general hospital). There are 5,588 EINs in total in this list.

I query more detailed information from the API on each hospital EIN. I save the name, secondary name, state, and zip code, none of which vary by year. I also record URL links to the Tax Form 990 PDFs, which are unique in each year. For the sake of a comprehensive data set, I keep years 2006-2020 (I later limit to 2010-2014 when focusing on pay-for-performance initiatives in 2012). Thus, I finish this step with a panel data set of EIN characteristics and PDF locations. Importantly, there are multiple types of Tax Form 990s depending on the size of the nonprofit. In many cases, one nonprofit has at least two different forms filed in a given year. I filter out any EIN-years for which there are no PDFs available. There are numerous foundations or auxiliary firms with the purpose of raising funds for the hospital, but do not provide services to patients. I filter out any not-for-profit with ``foundation" or ``auxiliary" in the name. I also filter out various specialty centers such as hospice or cancer centers.

@@

I proceed matching based on names in several steps. I focus on exact string matches, so I remove all spaces and common characters that could cause mismatches such as \&, ', -, and ``inc". I loop through each AHA ID, limit to nonprofits in the same state, and extract any exact name matches. When an exact match is found, I record the link between AHA ID and EIN. In this first layer of matching, 860 hospitals in the AHA data are linked to an EIN, equivalent to 31\% of nonprofit hospitals in the sample. In the next layer of matching, I remove common words such as ``healthcare", ``regional", ``hospital", etc. That way if there are subtle differences in names, removing common words may allow for an exact match. Again, I take each AHA hospital name and look for exact matches in the nonprofits in the same state. This adds an additional 90 hospital matches, accounting for a total of 34.5\% of AHA hospitals. Finally, I manually search through unmatched EINs to identify any matches. From Google searches, I identify an additional 300 EIN-AHA ID matches. yielding a total of 1,200 EINs.

Although analyzing all nonprofit hospitals would be ideal, a sample size of 852 firms is considered relatively large in the executive literature. I present a comparison of in-sample and out-of-sample nonprofits. I present means and standard deviations of these measures in Table \ref{tab:nfp_sample_compare}. The in-sample nonprofits are slightly larger in terms of beds and patients seen. They are also slightly more likely to be penalized or receive payments under the pay-for-performance incentives. However, average readmission and mortality rates are very similar, as well as case mix index. The largest difference in the two samples is that I under-represent nonprofit hospitals in systems. This is due to the nature of the tax form 990s, where systems often file one tax form for the entire system, making it difficult to discern the true managers of a specific hospital. Because of this, I drop hospitals from the sample who only have leadership information from a system tax form.

@@

I then extract the names of board members and executives from the Tax Form 990 PDFs for matched hospitals. In the data set of hospital PDF URLs that I collected earlier, I limit to the hospitals with solid matches described above. I then loop through each EIN, downloading PDFs locally and using the tesseract package in R to extract text from the relevant pages of the PDF using OCR text extraction methods. In particular, I loop through each page of the PDF, look for the title associated with leadership names: ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees", and save all the text from any pages where this title is found. I save the text to a list of all EIN, years present.

One aspect of the NonProfit Explorer API is that in some cases, if two forms are present for an EIN-year, only the first one is pulled. Therefore, for some hospitals, a couple years will have gaps in text extraction data. I locate EIN-years where this problem is occurring, and a team of RAs locates and downloads the correct forms manually. I extract text from these manually downloaded forms in the same manner as above.

@@

The data is now an executive-year-level data set with information on clinical training and specialty when relevant. I then create hospital-level indicators based on the names, titles and positions associated with the hospital: the number of clinically trained executives, the number of executives with particular specialties, the number of total executives, and whether the hospital employs a CMO.

@@

For the main analysis, I limit to hospitals who either always have a clinically trained executive or never have a clinically trained executive. In a supplemental analyses, I leverage changes to hospital propensity to hire a clinically trained executive. I explore the exogeneity of such changes in Section \ref{sec:changes}. Additionally, I present the frequency of hospitals who change their propensity to hire a clinically trained executive at different times in Table \ref{tab:change_timing}. I break the timing up into pre-2012, 2012, and post-2012. I show the number of hospitals who have clinically trained executives in different time combinations. The number of hospitals that only have an MD executive in 2012, but at no other time, is only 6. The majority of hospitals (197) who ever have a clinical executive have one in every time period.

@@

In this section, I present several additional figures and tables providing more information on the data. In Figure \ref{fig:state_doc}, I show that clinical executives are not concentrated in one region geographically. In Table \ref{tab:fp_samples_stable}, I provide means of all relevant variables for for-profit hospitals. In Table \ref{tab:specialty}, I show means of all relevant variables for hospital with and without internal medicine executives. In Figures \ref{fig:main_read_control_weights} and \ref{fig:main_read_time_weights}, I provide the time and control weights from the synthetic difference in differences estimation with readmission rate as the outcome. In Figures \ref{fig:main_mort_control_weights} and \ref{fig:main_mort_time_weights}, I provide the time and control weights from the synthetic difference in differences estimation with mortality rate as the outcome.

@@

    \subsection{Intensive Margin, Controlling for Num. Executives}\label{app:binned}

@@
